### 2017-05-06 2017mergingmerges/section/01-Introduction.tex
serious cutting
Make intro fit on the first page

We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream. Each operator could be given this interface, so that values from each stream are computed on demand. This approach is commonly taken in database systems \cite{Graefe:Volcano}. However, this `pull' model does not support operators with multiple outputs, such as our example, at least not without unbounded buffering. Suppose a consumer pulls many elements from the @sUnique@ output stream. In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering an arbitrary number of them. It needs to buffer these elements because they are also needed to perform the @merge@ operation. When a consumer finally pulls from @sUnion@ we will be able to drain the buffer of @sIn1@ elements, but not before.

Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the implied dataflow network and produces a simple sequential loop that does not require extra process-control abstractions or unbounded buffering. Sadly, existing stream fusion transformations cannot handle our example. As observed by \citet{kay2009you}, both pull-based and push-based fusion have fundamental limitations. Pull-based systems such as short-cut stream fusion~\cite{coutts2007stream} cannot handle cases where a particular stream or intermediate result is used by multiple consumers. We refer to this situation as a \mbox{\emph{split} --- in the} dataflow network of our example the flow from input stream @sIn1@ is split into both the @group@ and @merge@ consumers.

Push-based systems such as foldr/build fusion~\cite{gill1993short} cannot fuse our example either, because they do not support operators with multiple inputs. We refer to this as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow network. Some systems support both pull and push: data flow inspired array fusion using series expressions~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators. More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.

Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.

@@

In summary, we make the following contributions:
\begin{itemize}
\item a process calculus for infinite streaming programs (\S\ref{s:Processes});
\item a fusion algorithm, the first to support splits and joins (\S\ref{s:Fusion});
\item benchmarks showing significant performance gains (\S\ref{s:Benchmarks});
\item proof of soundness for the fusion algorithm in Coq (\S\ref{s:Proofs}).
\end{itemize}

>>>>

% We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream. Each operator could be given this interface, so that values from each stream are computed on demand. This approach is commonly taken in database systems \cite{Graefe:Volcano}. However, this `pull' model does not support operators with multiple outputs, such as our example, at least not without unbounded buffering. Suppose a consumer pulls many elements from the @sUnique@ output stream. In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering an arbitrary number of them. It needs to buffer these elements because they are also needed to perform the @merge@ operation. When a consumer finally pulls from @sUnion@ we will be able to drain the buffer of @sIn1@ elements, but not before.

Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the implied dataflow network and produces a simple sequential loop that does not require extra process-control abstractions or unbounded buffering. Sadly, existing stream fusion transformations cannot handle our example.

As observed by \citet{kay2009you}, both pull-based and push-based fusion have fundamental limitations. Pull-based systems such as short-cut stream fusion~\cite{coutts2007stream} cannot handle cases where a particular stream or intermediate result is used by multiple consumers. We refer to this situation as a \mbox{\emph{split} --- in the} dataflow network of our example the flow from input stream @sIn1@ is split into both the @group@ and @merge@ consumers. Push-based systems such as foldr/build fusion~\cite{gill1993short} cannot fuse our example either, because they do not support operators with multiple inputs. We refer to this as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow network. Some systems support both pull and push: data flow inspired array fusion using series expressions~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators. More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.

% Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.

% The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.

@@

In summary, we make the following contributions:
\begin{itemize}
\item a process calculus for infinite streaming programs (\S\ref{s:Processes});
\item a fusion algorithm, the first to support splits and joins (\S\ref{s:Fusion});
\item benchmarks showing significant performance gains (\S\ref{s:Evaluation});
\item proof of soundness for the fusion algorithm in Coq (\S\ref{s:Proofs}).
\end{itemize}
